---
title: "Keystone"
author: "Paule Mathieu"
format: html
editor: visual
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

## Introduction

You have inherited a project from a former PhD student who was studying the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster.

They left you with four separate datasets:

1.  estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs.
2.  estuary_metadata.csv: The spatial coordinates and estuary zones for each site.
3.  estuary_sonde_data.csv: High-frequency water quality sensor readings.
4.  species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.

Your mission is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

### Clear out environment

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

### Load packages

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readxl)
```

### Load files
```{r}
metadata <- read_csv("F:/Paule/MB5370/MB5370_Module_02/data/workshop2/estuary_metadata.csv")
```

```{r}
file_path <- ("F:/Paule/MB5370/MB5370_Module_02/data/workshop2/estuary_catch_log.xlsx")
catch_data2 <- excel_sheets(file_path) %>%
  map_dfr(~ read_excel(file_path, sheet = .x))
```
Note: So this is the wrong path? Because someone else could not access my data?
