---
title: "Workshop 02"
author: "Paule Mathieu"
format: html
editor: source
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Workshop 02 - Advanced data wrangling (extracting ecological signals from noisy systems)

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

## Introduction

This workshop of *R for Marine Science* builds on functional programming and data wrangling skills, focusing on extracting meaningful biological patterns from complex ecological datasets.

### Clear out environment

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

### Load packages

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

### 2.2 Tidying data

How we make our dataset tidy is by following three interrelated rules:

\- Each variable must have its own column.

\- Each observation must have its own row.

\- Each value must have its own cell.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Visualize changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

### 2.3 Lengthening datasets

```{r}
#What does my data look like?
billboard
#We've got a tibble of 317 rows and 79 columns, where each row represents a song and each column represents a week on the chart. The first 7 columns contain metadata about the songs, and the remaining columns contain the rank of the song for each week.

#We want to pivot because actual data values (wk1, wk2 etc.) are in the column name, with each observation (row of data) being a song. We need to have the data in a format where each row is an observation (long format).

#Pivoting the data
billboard %>%
  pivot_longer(
    cols = starts_with("wk"), # Specify the columns to pivot
    names_to = "week",        # Name of the new column for the week variable
    values_to = "rank"       # Name of the new column for the rank variable
  ) %>%
  filter(!is.na(rank))       # Remove rows where rank is NA OR values_drop_na = TRUE

```

```{r}
#Pivoting longer
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

#We want our new tidy dataset to have three variables: id, measurement (column name), and value (cell values).
df %>%
  pivot_longer(
    cols = bp1:bp2, # Specify the columns to pivot
    names_to = "measurement", # Name of the new column for the measurement variable
    values_to = "value"      # Name of the new column for the value variable
  )

#Note: I have definitely done this before but in excel.
```

### 2.5 Widening datasets

```{r}
cms_patient_experience

#We want to pivot wider because the data is in a long format where each row represents a patient experience, and we want to reshape it so that each row represents a unique patient with separate columns for each experience type.
cms_patient_experience %>%
  distinct(measure_cd, measure_title)

cms_patient_experience %>%
  pivot_wider(
    names_from = measure_cd, # Specify the column to use for new column names
    values_from = prf_rate   # Specify the column to use for new column values
  )

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  ) #this looks ugly though... like the output is not visually pleasing
```

```{r}
#Pivoting wider
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df %>%
  pivot_wider(
    names_from = measurement, # Specify the column to use for new column names
    values_from = value      # Specify the column to use for new column values
  )

df |> 
  distinct(measurement) |> 
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

#Combining both to create an empty data frame
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
#Not sure what the goal of creating an empty dataset does, since we already succeeded in widening our dataset using pivot_wider()...

```

### 2.6 Pivoting exercises using palmer penguins

```{r}
library(palmerpenguins)
#Create long version of penguins dataset
penguins_long <- penguins %>%
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
    names_to = "measurement_type", 
    values_to = "value"
  )
#View result
head(penguins_long)

#Visualizing
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```

```{r}
#Calculate the mean body mass for each species on each island using group_by() and summarise().
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

#Create wide version of penguins dataset
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,    #dictates which column provides our new column headers (the islands)
    values_from = mean_mass #dictates which column populates the cells beneath them (the mean mass)
  )

head(mass_matrix)

```

### 2.7 Separating and uniting data tables

```{r}
table3
#Separate the rate column into cases and population using separate().
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/") #specify the separator as "/" since the rate column contains values in the format "cases/population"

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) #By setting convert = TRUE, we can automatically convert the separated values into appropriate data types (e.g., numeric) instead of keeping them as character strings.

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) #This makes this data less tidy. but it is an example of how to separate a column into two parts based on a specific position (after the first two characters) rather than a separator. The century column will contain the first two characters of the year, and the year column will contain the remaining characters. Which can be useful for certain analyses where you want to analyze trends by century or decade, but it does make the data less tidy since we now have two columns representing a single variable (year).

```

```{r}
#To perform the inverse of separate() we will use unite() to combine multiple columns into a single column. In the example below for table5, we use unite() to rejoin century and year columns. unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select(). 

table5 %>% 
  unite(new, century, year, sep = "") #This combines the century and year columns into a single year column without any separator (sep = ""). The resulting year column will contain values like "1999" or "2000", which is more tidy and easier to work with for analyses that require a single year variable.
```

### 2.8 Wrangling strings and dates

```{r}
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c("Nelly Bay", "nelly_bay", "NELLY BAY", "Geoffrey_Bay", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
    # 3. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean)
  )

print(clean_sites)

```

```{r}
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)

#This allows us to perform date-time operations, such as calculating time differences, extracting components (like year, month, day), and plotting time series data accurately.

```

### 2.9 Relational data: Joining tables
```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)


# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "General Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

#Joining metadata to our observations
joined_data <- observations %>%
  left_join(site_metadata, by = join_by(site_code))
```

```{r}
#An inner_join() only keeps rows that have complete matches in both tables. If an observation occurs at a site that isn't listed in the metadata, that observation will be entirely dropped from the dataset. Use this when you only want to analyze data that has complete spatial or environmental context.
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)

# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code)) #anti_join() returns all rows from the first table (observations) that do not have a match in the second table (site_metadata). This allows us to identify which observations are missing spatial context and may need to be excluded from certain analyses or require additional data collection.) 

```

### 2.10 Handling missing values
```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

#This allows us to perform accurate calculations and analyses without the influence of erroneous values. We can now calculate summary statistics, create visualizations, or apply models to the temperature data without worrying about the -999 values skewing our results.
```

```{r}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)


```
Bonus tip: coalesce() is incredibly powerful. If you have two columns, like a primary_sensor and a backup_sensor, coalesce(primary_sensor, backup_sensor) will automatically use the primary reading, but seamlessly fill in the gaps with the backup reading if the primary had failed!

```{r}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)

#if you need to specifically audit your data for mathematical errors without flagging standard missing data, you can isolate them using the is.nan() function.

```

```{r}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))

# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

```
```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)

print(clean_log)
```

### 2.11 Practical exercises: Penguins, pivots and relational data

#### Exercise 1: Cleaning the messy metadata
```{r}
# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)
```
```{r}
#Task: Create a new object called clean_metadata. Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset (hint: watch out for those hidden spaces!). Then, convert the station_install column into a proper Date object.
clean_metadata <- island_metadata |>
  mutate(
    island_name = str_trim(island_name), # Remove leading and trailing whitespace
    station_install = dmy(station_install) # Convert to Date format
  )
print(clean_metadata)
```

#### Exercise 2: The relational join
```{r}
#Task: Create a new object called penguins_spatial. Use a left_join() to attach your clean_metadata to the main penguins dataset. Hint: Because the column containing the island names is called island in the penguins dataset, and island_name in your metadata, you will need to specify how they match using by = join_by(island == island_name).
penguins_spatial <- penguins |>
  left_join(clean_metadata, by = join_by(island == island_name))
print(penguins_spatial)
head(penguins_spatial)
```

#### Exercise 3: The wide summary matrix
```{r}
#Starting with your penguins_spatial dataset:
  #Drop any missing values in the body_mass_g column.
  #Group by species and island.
  #Summarise the data to find the maximum body mass (max()).
  #Pivot the dataset wider so that the species form the rows, and the islands form the columns.
penguins_summary <- penguins_spatial |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(max_mass = max(body_mass_g)) |>
  pivot_wider(names_from = island, values_from = max_mass)
print(penguins_summary)
head(penguins_summary)
```


