---
title: "Workshop 02"
author: "Paule Mathieu"
format: html
editor: source
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Workshop 02 - Advanced data wrangling (extracting ecological signals from noisy systems)

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

## Introduction

This workshop of *R for Marine Science* builds on functional programming and data wrangling skills, focusing on extracting meaningful biological patterns from complex ecological datasets.

### Clear out environment

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

### Load packages

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

### 2.2 Tidying data

How we make our dataset tidy is by following three interrelated rules:

\- Each variable must have its own column.

\- Each observation must have its own row.

\- Each value must have its own cell.

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Visualize changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

### 2.3 Lengthening datasets
```{r}
#What does my data look like?
billboard
#We've got a tibble of 317 rows and 79 columns, where each row represents a song and each column represents a week on the chart. The first 7 columns contain metadata about the songs, and the remaining columns contain the rank of the song for each week.

#We want to pivot because actual data values (wk1, wk2 etc.) are in the column name, with each observation (row of data) being a song. We need to have the data in a format where each row is an observation (long format).

#Pivoting the data
billboard %>%
  pivot_longer(
    cols = starts_with("wk"), # Specify the columns to pivot
    names_to = "week",        # Name of the new column for the week variable
    values_to = "rank"       # Name of the new column for the rank variable
  ) %>%
  filter(!is.na(rank))       # Remove rows where rank is NA OR values_drop_na = TRUE

```

```{r}
#Pivoting longer
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

#We want our new tidy dataset to have three variables: id, measurement (column name), and value (cell values).
df %>%
  pivot_longer(
    cols = bp1:bp2, # Specify the columns to pivot
    names_to = "measurement", # Name of the new column for the measurement variable
    values_to = "value"      # Name of the new column for the value variable
  )

#Note: I have definitely done this before but in excel.
```

### 2.5 Widening datasets
```{r}
cms_patient_experience

#We want to pivot wider because the data is in a long format where each row represents a patient experience, and we want to reshape it so that each row represents a unique patient with separate columns for each experience type.
cms_patient_experience %>%
  distinct(measure_cd, measure_title)

cms_patient_experience %>%
  pivot_wider(
    names_from = measure_cd, # Specify the column to use for new column names
    values_from = prf_rate   # Specify the column to use for new column values
  )

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  ) #this looks ugly though... like the output is not visually pleasing
```

```{r}
#Pivoting wider
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df %>%
  pivot_wider(
    names_from = measurement, # Specify the column to use for new column names
    values_from = value      # Specify the column to use for new column values
  )

df |> 
  distinct(measurement) |> 
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

#Combining both to create an empty data frame
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
#Not sure what the goal of creating an empty dataset does, since we already succeeded in widening our dataset using pivot_wider()...

```

### 2.6 Pivoting exercises using palmer penguins


