---
title: "Workshops"
author: "Paule Mathieu"
date: "2026-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workshop 1 - Foundations of Data Science (Wrangling & Plotting) ###
```{r}
# Installing packages
library(tidyverse)
library(readxl)
#install.packages("here")
library(here)
```

```{r}
# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")
```

```{r}
# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/mangrove_survey_raw.csv"))

# Cleaning up a messy .csv file
# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

```
```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
```

```{r}
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)

#I can't really see the difference to be honest, feels very minor, but overall, should print and look at the data after manipulating it to make sure it is responding as it should, got it.
```
```{r}
# Install the data package
#install.packages("palmerpenguins")

# Load the package data into active memory
library(palmerpenguins)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

#I do love a good summary matrix
```
# Isolating attributes with select()
```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)
```

# Sifting rows with filter()
```{r}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

# Ordering sequences with arrange()
```{r}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
#I often use a combination of all three depending on what my data looks like, or I will preemptively format my datasheet in a way that makes it more easy to use in R, before I start collecting. But that is not always possible.
```

# Using the Pipe
```{r}
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")

penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
#I have definitely used this before, this seems more direct tbh.
```

# Computing new attributes with mutate()
```{r}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)
#I would still remove the NAs or replace them with 0 or something, cause it could cause problem in later statistical analysis.
```
```{r}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```
```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)
```
# I now realise that that I should probably be doing my data clean up inside a R script and just source that clean one into a markdown, which I will make sure to do next time but it's okay.

# Visualizing body size and shape
```{r}
# Summary table
penguins |>
  group_by(species, island) |>
  summarise(
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    .groups = "drop"
  )

# Visualisation
penguin_figure <- 
  penguins |>
  ggplot(aes(x = species, y = body_mass_g, fill = island)) +
  geom_boxplot() +
  scale_fill_manual(
    name = "Penguin Island", 
    values = c(
      "Biscoe" = "skyblue",
      "Dream" = "purple", 
      "Torgersen" = "navy"
    )
  ) +
  labs(
    title = "Body Mass by Species and Island",
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

# Piping directly into visualization
```{r}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
    scale_colour_manual(
    name = "Penguin Island",
    values = c(
      "Biscoe" = "skyblue",
      "Dream" = "purple",
      "Torgersen" = "navy"
    )
  ) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot
```


